The Magnetorotational Instability Prefers Three Dimensions

### 2019-11-12 paper/referee_response.tex

\documentclass[letterpaper, 12pt]{article}

% Imports
\usepackage{amsmath, amssymb}
\usepackage{mathtools}
\usepackage{graphicx}

% Layout
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

% Body font
\usepackage{soul, color}
\usepackage[colorlinks, allcolors=blue]{hyperref}
\usepackage{newtxtext}

% Reviews
\usepackage{framed}
\newenvironment{review}{\vspace{1em}\begin{leftbar}}{\end{leftbar}}

\begin{document}

Dear Professor Tobias,

We thank the referees for constructive reports. We have made all of the changes they suggested. In addition, we added section 3c detailing an asymptotic confirmation of the 3D to 2D transition reported numerically in section 3a to at least three significant figures. If this section is too much of an addition at this stage, we can relegate it to an appendix or remove it.

Please see our responses in line below.

\begin{review}
Comments to the Author(s)
This is an interesting manuscript that reveals a new and important result regarding the magnetorotational instability (MRI) in a radially confined geometry such as a Taylor-Couette experiment between differentially rotating cylinders. The authors show that 3D (non-axisymmetric) modes can be preferred to 2D (axisymmetric) ones, especially near the onset of the 2D instability. This might be important for the outcome of MRI experiments such as the one at Princeton, and it also has ramifications for dynamo action. The authors obtain good agreement, where expected between their numerical calculations of unstable modes and their asymptotic analysis.

The manuscript is mostly well written and the figures are well presented (although some are too small in the current format - in particular, Fig. 3 is hard to read).

I recommend publication in Proceedings A provided the following matters can be addressed.

1. It would be helpful to explain that "three-dimensional" means "non-axisymmetric" and "two-dimensional" means "axisymmetric" (but with three-component velocity and magnetic fields). (Of course, "axisymmetric" is to be interpreted as "y-independent" in this local model.)
\end{review}
We have fixed this.

\begin{review}  
2. Section 1: Perhaps clarify that the "finite-channel cutoff" is due to the finite width, not height.
\end{review}
Fixed.

\begin{review}
3. Section 1: What does "weak-shear" mean here: weak compared to what?  
\end{review}

Fixed

\begin{review}
4. Section 2: The notation could be defined more clearly by saying what v, b and j represent. We certainly need to know that f = 2 Omega.
\end{review}

We are unclear what the referee refers to here; these are already in the text.

\begin{review}
5. Section 2: The channel is periodic in z, but it seems that the periodicity length is unspecified, so that any value of the wavenumber $k_z$ is permissible. Please clarify this.
\end{review}

Fixed.

\begin{review}
6. Section 2: Personally, I find it confusing to have the symbol q denoting the Rossby number when (as said in the footnote) in most of the MRI literature q denotes twice the Rossby number. Would it not be better to use a different symbol for the Rossby number?
\end{review}

Fixed

\begin{review}
7. Section 2: I believe that q > 1 (rather than q = 1, as stated) corresponds to Rayleigh-unstable shear. (Are we to assume throughout that q > 0, i.e. S < 0?)
\end{review}
Fixed

\begin{review}
8. Around eq. 2.6: Please clarify the meaning of "in two dimensions" (i.e. axisymmetric, or y-independent, perturbations). Is there a convenient reference for eq. 2.6, as the result is not obvious? I believe it comes from taking the limit $k_z d \ll 1$, so the periodicity length in z has to be very long.
\end{review}

Fixed.

\begin{review}
9. Section 2: An important point is that the system of units has not been clarified and the relevant dimensionless parameters have not all been identified. Everything is expressed in dimensional terms until we are told that nu = eta = 1e-5. In what units are these values expressed?
  \end{review}

We have clarified this point while simultaneously addressing point 16.

\begin{review}
10. Section 2: I thought initially that there was a missing, and unspecified, dimensionless parameter such as $B_0 / f d$, which is a measure of the field strength. I see now that this can be determined from the given values of $q$ and $S/S_c$, but it might be better to explain this in the text.
\end{review}

We have done so.

\begin{review}
11. Figure 1 caption: Presumably "cray" should be "grey".  
\end{review}

Fixed

\begin{review}
12. Before eq. 3.1: Presumably "in when" should be "when".  
\end{review}

Fixed

\begin{review}
13. Before eq. 3.9: "2nd order" should probably be "2nd order in x".  
\end{review}

Fixed

\begin{review}
14. Eq. 3.9: Why are $k_z$ and $k_x$ renamed $k$ and $l$ here?  
\end{review}

We have corrected this.

\begin{review}
15. Section 3(c): It seems to me that the prevalence of axisymmetric modes in previous numerical simulations may have more to do with the radial boundary conditions than with the degree of supercriticality.
\end{review}

A note was added to this effect.

\begin{review}
16. Section 3(e): It would have been better to define Re and Rm back in Section 2. Again, the units in which eta and nu are given are not specified. There are a couple of typos in the first sentence: "as series" should be "a series", and "with the" should be "with".
\end{review}

Typos fixed and Reynolds numbers moved upwards.

\begin{review}
17. I disagree that there is only one previous study of non-turbulent MRI dynamos; indeed, reference [15] and related papers provide examples of this.  
\end{review}

This is completely correct; our original paper was unclear. We meant to express that there are no other *linear* MRI dynamos, not *laminar* ones!

\begin{review}
Referee: 2

Comments to the Author(s)
Thank you very much for sending your article! I find the subject, three-dimensional (3d) modes of the magneto-rotational instability (MRI), a very interesting aspect that so far has received much less attention than the corresponding two-dimensional (2d) modes. Your analysis is very well suited to the task, the results are interesting and presented clearly and in detail and set into the context of conditions in stars and lab experiments . Hence I recommend the article for publication. I have, however, a few minor points to
raise. The primary one is a question regarding the range of applicability of the 3d results, in particular in stars. You vary the ratio between the shear, S, and the critical shear value, $S_c$, given in equation (2.6), from values below unity to a maximum of 4. You note that for $S/S_c >~ 2.05$, the linear dynamics should be dominated by 2d
modes. $S_c$ involves, among other quantities, the square of the magnetic field strength, which can vary by (or is uncertain to) quite large factors. Thus, even moderate variations of the field may change $S_c$ such that $S/S_c$ no longer favours 3d modes. My question is, thus, if this apparent need for what might be considered fine-tuning limits the applicability of the results to, e.g., stars.
\end{review}
Thank you for pointing this out. We have added a cautionary note regarding this point.

\begin{review}
Apart from that, I have only small points:
- Section 3(a): the text mentions $S/S_c = 1.02$, but Fig. 1 has 1.002 and 1.2. Also, at the end of the paragraph: "does not intercept the x-axis"; one might put "$k_z$-axis" to avoid confusion with the axes of the simulation box.
- Section 3(f), after the equation: "rms is denotes...": remove one of the two verbs.
- Also 3(f): does the integral of the EMF over x vanish for reasons of symmetry?
\end{review}

We have fixed all of these.

\vspace{1em}

Sincerely, \\
Jeffrey S. Oishi, Geoffrey M. Vasil, Morgan Baxter, Andrew Swan, Keaton J. Burns, Daniel Lecoanet, \& Benjamin P. Brown
\end{document}

%%% Local Variables:
%%% ispell-local-dictionary: "british"
%%% End:

%  LocalWords:  Tobias magnetorotational Couette axisymmetric Rossby
%  LocalWords:  wavenumber eq nd supercriticality rms Oishi Vasil
%  LocalWords:  Lecoanet

### 2019-11-12 paper/paper.tex
checked all new equations. did calc. from start. everything checks out.
Final revisions and response to referees.

In order to ensure our results are converged, we have repeated runs at $n_x=256$ Chebyshev modes as well as doubling the number of modes in $y$ and $z$.

\begin{tabular}{cccccccccccccc}
\textbf{Run} & \textbf{$\SSC$} & \textbf{$\nu$} & \textbf{$\eta$} & \textbf{$\Ro$} & \textbf{$N_x$} & \textbf{$N_{k_{y}}$} & \textbf{$N_{k_{z}}$} & \textbf{Sparse/Dense}& \\
1  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & resolution study\\
2  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 256 & 200 & 200 & sparse & \\
3  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & dense  & \\
4  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 512 & 512 & sparse & \\
5  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 100 & 100 & sparse & \\
6  &   0.2  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & $\SSC$ variation\\
7  &   0.3  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
8  &   0.4  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
9  &   0.5  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
10 &   0.64 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
11 & 1.002  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
12 & 1.002  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & dense & \\
13 & 1.44   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
14 & 1.75   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
15 & 1.891  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
16 & 2.     & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
17 & 2.01   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
18 & 2.015  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
19 & 2.031  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
20 & 2.05   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
21 & 2.1    & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
22 & 2.25   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
23 & 2.5    & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
24 & 4      & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
25 & 1.02   & $10^{-6}$ & $10^{-6}$ & 0.75 & 128 & 200 & 200 & sparse & Reynolds number study\\
26 & 1.02   & $10^{-4}$ & $10^{-4}$ & 0.75 & 128 & 200 & 200 & sparse & \\
27 & 1.02   & $10^{-3}$ & $10^{-3}$ & 0.75 & 128 & 200 & 200 & sparse & \\
28 & 1.02   & $10^{-2}$ & $10^{-2}$ & 0.75 & 128 & 200 & 200 & sparse & \\
29 & 1.02   & $10^{-5}$ & $10^{-5}$ & 0.1  & 128 & 200 & 200 & sparse & Low Rossby\\ 
30 & 1.02   & $10^{-6}$ & $10^{-2}$ & 0.1  & 128 & 200 & 200 & sparse & Liquid metal case\\ 
\end{tabular}
%\label{simulations}

@@

We can understand the transition from three- to two-dimensional instability semi-analytically. We do this by conducting a similar analysis as the above asymptotic calculation, only close to the fastest growing 2D modes. We can look for values for $\SSC$ where the analogous value of $\Upsilon(\SSC) = 0$.

We analyse the growth rate as a function of  $k_{y},k_{z}$ and $\SSC$ near $k_{y} \approx0$.  We therefore expand the whole problem in a power-series ansatz
\Beq
\gamma(\SSC,k_{y},k_{z}) \ = \   \sum_{n\ge 0} \gamma_{n}(\SSC,k_{z}) k_{y}^{n}, \quad \psi(x) \ = \ \sum_{n\ge 0} \psi_{n}(x) k_{y}^{n}
\Eeq
We use the same asymptotic techniques as before to obtain each progressive correction. 
\Beq
\psi_{0}'' - \Phi_{0} \psi_{0} & = & 0 \label{eq:psi0} \\
\psi_{1}'' - \Phi_{0} \psi_{1} & = & \Phi_{1} \psi_{0} \label{eq:psi1} \\
\psi_{2}'' - \Phi_{0} \psi_{2} & = & \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1} \label{eq:psi2}  
\Eeq
We solve the leading order by virtue that all non-constant coefficients vanish for 2D modes. The leading-order eigenfunction is $\psi_{0} = \cos(\pi x / d)$, just as before. The following dispersion relations determine the maximum growth rate and the corresponding wavenumber,
\Beq
\mathfrak{D}_{0}(k_{z}) &=& \frac{f k_z^2 \left(B_0^2 S k_z^2+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_z^2+\gamma _0^2\right){}^2}+\frac{\pi ^2}{d^2}+k_z^2 \ = \ 0 \\
\frac{\mathfrak{D}_{0}'(k_{z})}{2k_{z}} &=& \frac{\gamma _0^2 f \left(B_0^2 k_{z}^2 (S-f)+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_{z}^2+\gamma _0^2\right)^3}+1 \ = \ 0.
\Eeq
We transform these relations into polynomials in terms of non-dimensional variables,
\Beq
\label{dispersion}
\y^2 \s^2 \Ro^2 +  \y \, \s  \, ( \y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s ) \x + ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x^2 +\x^3 &=& 0 \\ 
\y \, \s  \, (\y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s    )  + 2 ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x +3\x^2 &=& 0,
\label{critical-dispersion}
\Eeq
where
\Beq
\x  \ = \ \frac{d^2 k_z^2}{\pi ^2}, \quad  \quad \y \ = \ \frac{\gamma_{0}^{2}}{S^{2}}, \quad \quad \s  \ = \ \frac{S}{S_{c}}.
\Eeq
These are directly related to the parameters reported on figure~\ref{fig:growth_rate}.
The two polynomials are straightforward to solve numerically for $\x,\y$ given $\s, \Ro$. However we  still want to find critical values for $\s$, where the fastest growing modes are two dimensional. From now on, the calculation becomes sufficiently complicated that we only outline the methodology.

We find the higher-order $\gamma_{n}(\s,k_{z})$ coefficients by applying order-by-order solvability conditions. We first find that $\gamma_{1}(\s, k_{z}) = 0$ identically. This renders the first-order correction \eq{eq:psi1} solvable. The 1st-order eigenfunction takes the form 
\Beq
\psi_{1}(x) \propto F(\pi x/d) \quad \text{where} \quad  F(\vartheta) = \vartheta (\vartheta \sin (\vartheta)+\cos (\vartheta))-\frac{\pi^{2}}{4} \sin (\vartheta).
\Eeq
The structure of the growth rate is therefore 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2} \ + \ \ldots 
\Eeq
This is the same structure as the asymptotic calculation in the previous section. Now effectively $\Upsilon \propto \gamma_{2}$. The case with dominant 2D modes corresponds to $\gamma_{2}(\s,\x)=0$.  The final result derives from the 2nd-order solvability condition 
\Beq
\int_{-d/2}^{d/2} ( \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1} ) \psi_{0} \dd{x}  \ = \ 0,
\Eeq
which produces the degree-6 polynomial expression  
\Beq
& \y^{6} \s^6\Ro^6  \ + \
 \y^4 \s^5 \Ro^4   \left( (3 \xi  \s+6 \y-1)\Ro -3 \xi  \s \right) \, \x +
  \nonumber \\  
& \frac{ \y^3 \s^4 \Ro^2  }{4}  \left( (60 \y -4 \xi  (\s-8) \s-16 + 3 \s^2)\Ro^{2} + 2 \s (4 \xi  (\s+1)-3 \s) \Ro - (4 \xi -3 ) \s^2 \right) \, \x^{2} + \nonumber \\ 
& \frac{\y^2  \s^3 \Ro }{2}  \left( (3 \s^2-4 \xi  (\s-3) \s+40 \y-12) \Ro^{2} + 24 \xi \s \Ro + (4 \xi -3) s^2 \right) \, \x^{3} + 
\nonumber \\ 
&  \frac{\y \,\s^2}{4} \left( (60 \y -(4 \xi -3) \s^2 - 16)\Ro^{2} + 2\s(3 \s-4 \xi  (\s-3)) \Ro - (4 \xi -3) \s^2 \right)\, \x^{4} +\nonumber \\ 
 &  \s\, (( 6 \y -\xi  \s-1) \Ro- \xi  \s) \, \x^{5}  \ + \   
   \x^{6} \ = \ 0.
   \label{2D-condition}
\Eeq
The new coefficient is
\Beq
\xi \ = \ \frac{\pi^{2}-6}{12} \ = \ \frac{\int_{-d/2}^{d/2} x^2 \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}{\frac{d^{2}}{\pi^{2}}\int_{-d/2}^{d/2}  \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}.
\Eeq
This results from projecting the squared shear onto the leading-order eigenmodes, just as in the previous section.

Given $\Ro$, we solve \eq{dispersion}, \eq{critical-dispersion}, and  \eq{2D-condition} for $\x,\y,\s$ using Newton's method.

\begin{center}
\begin{tabular}{cccccccccccccc}
\textbf{$\Ro$} & \textbf{$k_{z} d/\pi$}  & \textbf{$\gamma_{\max}/|S_{2D}|$}  &  \textbf{$S_{2D}/S_{c}$} \\
                   1.00  & 0.689610 & 0.208727 &   2.09694 \\ 
                   0.75 & 0.682819 & 0.214895 & 2.05136 \\
                   0.50 & 0.676062 & 0.221309 & 2.00061 \\
                   0.25 & 0.671528 & 0.228551 & 1.94963 \\
                   0.10 & 0.673164 & 0.234376 & 1.92692 \\
                   0.00 & 0.462324 & 0.240212 & 1.92465
\end{tabular}
\end{center}
The numbers for $\Ro = 0.75$ match the numerical results shown in Figure~\ref{fig:growth_rate} excellently.

Finally, we can continue with the calculation conceptually. Because of symmetry $\gamma_{3}(\s,\x) = 0$ identically. This means that to leading order 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2}  + \gamma_{4}(\s,\x) k_{y}^{4} \ + \ \ldots
\Eeq 
For $\SSC > S_{2D}/S_{c}$, $\gamma_{2} > 0$. This is sufficient to determine the most unstable 3D wavenumber near the transition to dominant 2D modes, 
\Beq
k_{y}^{2} \ \approx \ -\frac{\gamma _2}{2 \gamma_4} \  \propto \   S_{2D}/S_{c} - \SSC   . 
\Eeq 
Generically $\gamma_{4} \ne 0$. We assume that $\gamma_{4} < 0$, but we do not know for sure if this is the case. The calculation is considerably more challenging than anything so far. If there is a negative 4th-order feedback, then 
\Beq
k_{y} \ \sim \ \sqrt{S_{2D}/S_{c} - \SSC}, 
\Eeq
which is consistent with Figure~\ref{fig:growth_rate}d near the critical point. Furthermore, a scenario where $\gamma_{4} > 0$ would require behaviour that is inconsistent with Figure~\ref{fig:growth_rate}d.

>>>>

In order to ensure our results are converged, we have repeated runs at $n_x=256$ Chebyshev modes as well as doubling the number of modes in $y$ and $z$.
\begin{table}
\begin{tabular}{cccccccccccccc}
\textbf{Run} & \textbf{$\SSC$} & \textbf{$\nu$} & \textbf{$\eta$} & \textbf{$\Ro$} & \textbf{$N_x$} & \textbf{$N_{k_{y}}$} & \textbf{$N_{k_{z}}$} & \textbf{Sparse/Dense}& \\
1  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & resolution study\\
2  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 256 & 200 & 200 & sparse & \\
3  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & dense  & \\
4  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 512 & 512 & sparse & \\
5  &   1.02 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 100 & 100 & sparse & \\
6  &   0.2  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & $\SSC$ variation\\
7  &   0.3  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
8  &   0.4  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
9  &   0.5  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
10 &   0.64 & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
11 & 1.002  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
12 & 1.002  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & dense & \\
13 & 1.44   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
14 & 1.75   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
15 & 1.891  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
16 & 2.     & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
17 & 2.01   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
18 & 2.015  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
19 & 2.031  & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
20 & 2.05   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
21 & 2.1    & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
22 & 2.25   & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
23 & 2.5    & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
24 & 4      & $10^{-5}$ & $10^{-5}$ & 0.75 & 128 & 200 & 200 & sparse & \\
25 & 1.02   & $10^{-6}$ & $10^{-6}$ & 0.75 & 128 & 200 & 200 & sparse & Reynolds number study\\
26 & 1.02   & $10^{-4}$ & $10^{-4}$ & 0.75 & 128 & 200 & 200 & sparse & \\
27 & 1.02   & $10^{-3}$ & $10^{-3}$ & 0.75 & 128 & 200 & 200 & sparse & \\
28 & 1.02   & $10^{-2}$ & $10^{-2}$ & 0.75 & 128 & 200 & 200 & sparse & \\
29 & 1.02   & $10^{-5}$ & $10^{-5}$ & 0.1  & 128 & 200 & 200 & sparse & Low Rossby\\ 
30 & 1.02   & $10^{-6}$ & $10^{-2}$ & 0.1  & 128 & 200 & 200 & sparse & Liquid metal case\\ 
\end{tabular}
\caption{Eigenvalue calculations performed.}
\label{tab:runs}
\end{table}
%\label{simulations}

@@

We can understand the transition from three- to two-dimensional instability semi-analytically. We do this by conducting a similar analysis as the asymptotic calculation in section~\ref{sec:asymp} near the fastest growing 2D modes, looking for values for $\SSC$ where the analogous value of $\Upsilon(\SSC)$ equals zero.

The growth rate can be written as a function of  $k_{y},k_{z}$ and $\SSC$ near $k_{y} \approx0$.  We therefore expand the whole problem in a power-series 
\Beq
\gamma(\SSC,k_{y},k_{z}) \ = \   \sum_{n\ge 0} \gamma_{n}(\SSC,k_{z}) k_{y}^{n}, \quad \psi(x) \ = \ \sum_{n\ge 0} \psi_{n}(x) k_{y}^{n},
\Eeq
and use the same asymptotic techniques as before to obtain each progressive correction,
\Beq
\psi_{0}'' - \Phi_{0} \psi_{0} & = & 0 \label{eq:psi0} \\
\psi_{1}'' - \Phi_{0} \psi_{1} & = & \Phi_{1} \psi_{0} \label{eq:psi1} \\
\psi_{2}'' - \Phi_{0} \psi_{2} & = & \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1}.
\label{eq:psi2}  
\Eeq
At leading order, we making use of the fact that all non-constant coefficients vanish for 2D modes. The leading-order eigenfunction is $\psi_{0} = \cos(\pi x / d)$, just as before. The following dispersion relations determine the maximum growth rate and the corresponding wavenumber,
\Beq
\mathfrak{D}_{0}(k_{z}) &=& \frac{f k_z^2 \left(B_0^2 S k_z^2+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_z^2+\gamma _0^2\right){}^2}+\frac{\pi ^2}{d^2}+k_z^2 \ = \ 0 \\
\frac{\mathfrak{D}_{0}'(k_{z})}{2k_{z}} &=& \frac{\gamma _0^2 f \left(B_0^2 k_{z}^2 (S-f)+\gamma _0^2 (f+S)\right)}{\left(B_0^2 k_{z}^2+\gamma _0^2\right)^3}+1 \ = \ 0.
\Eeq
We transform these relations into polynomials in terms of non-dimensional variables,
\Beq
\label{dispersion}
\y^2 \s^2 \Ro^2 +  \y \, \s  \, ( \y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s ) \x + ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x^2 +\x^3 &=& 0 \\ 
\y \, \s  \, (\y \, \s \, \Ro^2 + (2  - \s)  \, \Ro + \s    )  + 2 ( 2 \y \,\s \,  \Ro - \s  + 1 ) \x +3\x^2 &=& 0,
\label{critical-dispersion}
\Eeq
where
\Beq
\x  \ = \ \frac{d^2 k_z^2}{\pi ^2}, \quad  \quad \y \ = \ \frac{\gamma_{0}^{2}}{S^{2}}, \quad \quad \s  \ = \ \frac{S}{S_{c}}.
\Eeq
These can be computed from the parameters reported in figure~\ref{fig:growth_rate} and table~\ref{tab:runs}.
The two polynomials are straightforward to solve numerically for $\x,\y$ given $\s, \Ro$. However, we want to find critical values for $\s$, where the fastest growing modes are two dimensional.

We find the higher-order $\gamma_{n}(\s,k_{z})$ coefficients by applying order-by-order solvability conditions. We first find that $\gamma_{1}(\s, k_{z}) = 0$ identically. This renders the first-order correction \eq{eq:psi1} solvable. The 1st-order eigenfunction takes the form 
\Beq
\psi_{1}(x) \ = \ A_{1} F(\pi x/d) \quad \text{where} \quad  F(\vartheta) = \vartheta (\vartheta \sin (\vartheta)+\cos (\vartheta))-\frac{\pi^{2}}{4} \sin (\vartheta).
\Eeq
and
\Beq
A_{1} \ = \ \frac{i \gamma _0 d^3 f S k_z^2 \left(B_0^2 (f-S) k_z^2-\gamma _0^2
   (f+S)\right)}{2 \pi ^3 \left(B_0^2 k_z^2+\gamma _0^2\right){}^{\!3}}.
\Eeq
The structure of the growth rate is therefore 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2} \ + \ \ldots 
\Eeq
This is the same structure as the asymptotic calculation in section~\ref{sec:asymp}. Here, $\Upsilon \propto \gamma_{2}$ effectively. The case with dominant 2D modes corresponds to $\gamma_{2}(\s,\x)=0$.  The final result derives from the 2nd-order solvability condition 
\Beq
\int_{-d/2}^{d/2} ( \Phi_{2} \psi_{0} + \Phi_{1} \psi_{1} ) \psi_{0} \dd{x}  \ = \ 0,
\Eeq
which produces the degree-6 polynomial expression  
\Beq
& \y^{6} \s^6\Ro^6  \ + \
 \y^4 \s^5 \Ro^4   \left( (3 \xi  \s+6 \y-1)\Ro -3 \xi  \s \right) \, \x +
  \nonumber \\  
& \frac{ \y^3 \s^4 \Ro^2  }{4}  \left( (60 \y -4 \xi  (\s-8) \s -16 + 3 \s^2)\Ro^{2} + 2 \s (4 \xi  (\s+1)-3 \s) \Ro - (4 \xi -3 ) \s^2 \right) \, \x^{2} + \nonumber \\ 
& \frac{\y^2  \s^3 \Ro }{2}  \left( (3 \s^2-4 \xi  (\s-3) \s+40 \y-12) \Ro^{2} + 24 \xi \s \Ro + (4 \xi -3) s^2 \right) \, \x^{3} + 
\nonumber \\ 
&  \frac{\y \,\s^2}{4} \left( (60 \y -(4 \xi -3) \s^2 - 16)\Ro^{2} + 2\s(3 \s-4 \xi  (\s-3)) \Ro - (4 \xi -3) \s^2 \right)\, \x^{4} +\nonumber \\ 
 &  \s\, (( 6 \y -\xi  \s-1) \Ro- \xi  \s) \, \x^{5}  \ + \   
   \x^{6} \ = \ 0.
   \label{2D-condition}
\Eeq
The new coefficient is
\Beq
\xi \ = \ \frac{\pi^{2}-6}{12} \ = \ \frac{\int_{-d/2}^{d/2} x^2 \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}}{\frac{d^{2}}{\pi^{2}}\int_{-d/2}^{d/2}  \cos ^2\left(\frac{\pi  x}{d}\right) \dd{x}} \ \approx \ 0.322467.
\Eeq
This results from projecting the squared shear onto the leading-order eigenmodes, just as in the previous section.

We solve \eq{dispersion}, \eq{critical-dispersion}, and  \eq{2D-condition} for $\x,\y,\s$ using Newton's method for a variety of $\Ro$, yielding

\begin{center}
\begin{tabular}{cccccccccccccc}
\textbf{$\Ro$} & \textbf{$k_{z} d/\pi$}  & \textbf{$\gamma_{\max}/|S_{2D}|$}  &  \textbf{$S_{2D}/S_{c}$} \\
                   1.00  & 0.689610 & 0.208727 &   2.09694 \\ 
                   0.75 & 0.682819 & 0.214895 & 2.05136 \\
                   0.50 & 0.676062 & 0.221309 & 2.00061 \\
                   0.25 & 0.671528 & 0.228551 & 1.94963 \\
                   0.10 & 0.673164 & 0.234376 & 1.92692 \\
                   0.00 & 0.462324 & 0.240212 & 1.92465
\end{tabular}
\end{center}
The value of $\SSC = 2.05136$ at the 2D to 3D transition when $\Ro = 0.75$ matches the numerical results shown in Figure~\ref{fig:growth_rate} two all reported significant figures in the numerical calculation.

We conclude this section with some speculations concerning the higher order corrections. Because of symmetry, $\gamma_{3}(\s,\x) = 0$ identically. This means that to leading order 
\Beq
\gamma(\s,k_{y},k_{z}) \ = \ \gamma_{0}(\s,\x) + \gamma_{2}(\s,\x) k_{y}^{2}  + \gamma_{4}(\s,\x) k_{y}^{4} \ + \ \ldots
\Eeq 
For $\SSC > S_{2D}/S_{c}$, $\gamma_{2} > 0$. This is sufficient to determine the most unstable 3D wavenumber near the transition to dominant 2D modes, 
\Beq
k_{y}^{2} \ \approx \ -\frac{\gamma _2}{2 \gamma_4} \  \propto \   S_{2D}/S_{c} - \SSC   . 
\Eeq 
In general, $\gamma_{4} \ne 0$. We speculate that $\gamma_{4} < 0$, though the calculation is outside the scope of this work. If there is a negative 4th-order feedback, then 
\Beq
k_{y} \ \sim \ \sqrt{S_{2D}/S_{c} - \SSC}, 
\Eeq
which is consistent with Figure~\ref{fig:growth_rate}d near the critical point. However, a scenario where $\gamma_{4} > 0$ would require behaviour that is inconsistent with our numerical calculations.
